Computational methods in landscape ecology

### 2024-10-18 Sections/03-spatial-patterns.qmd

<!-- 
Software to calculate landscape metrics include the *landscapemetrics* [@Hesselbarth2019LandscapemetricsOpensource] and *multilandr* [@Huais2024MultilandrPackage] R packages, the *PyLandStats* Python library [@Bosch2019PyLandStatsOpensource], or the stand-alone software *FRAGSTATS* [@McGarigal2023FRAGSTATSV4].
-->

@@

<!-- 
Entropy metrics are available in various software such as the R packages *landscapemetrics*, *bespatial* [@Nowosad2024BespatialBoltzmann], *rasterdiv* [@Rocchini2021RasterdivInformation], *LandComp* [@Konrad2023AdaptingMultiscale], or the Python package *PyLandStats*.
-->

@@

<!--
Nonetheless, the *geodiv* R package [@Smith2021GeodivPackage] provides various surface metrics, and related textures from grey-level co-occurrence matrices are provided by the *glcm* R package [@Zvoleff2020GlcmCalculate].
Another approach to quantify the gradient surface model is based on frequencies and local adjacencies of continuous input pixel values as implemented in the *GraySpatCon* software [@Riitters2023MappingLandscape].
-->

## Vector-based metrics

@@

<!--
The *vectormetrics* R package [@Matuszek2024VectormetricsLandscape] is specifically designed for vector metrics.
However, all general software frameworks designed for spatial data handling which support vector data can theoretically be used to calculate basic vector metrics.
-->

@@

Spatial patterns can be analyzed through a range of computational operations, e.g., comparing, searching, or grouping.
These operations are based on spatial signatures (multi-numerical representations of landscape pattern) and dissimilarity measures (functions that quantify differences between the signatures), and can be calculated for different areas or for the same area at two different moments in time.
Comparing spatial signatures is often used to analyze landscape dynamics, e.g., to detect changes in landscape structure over time [@Duncan2023MonitoringSpatial].
<!-- R1: As defined here, it looks to me that this operation is not necessarily linked to a query area, but rather to a spatial signature which can be directly specified or extracted from a query area -->
Furthermore, signature-based searches can be used to compare the spatial signature of a focal area to the spatial signatures of multiple other areas.
This allows for the identification of areas with similar signatures compared to the focal area, e.g., areas with similar environmental conditions [@Jasiewicz2014LandscapeSimilarity].
Additionally, it is possible to calculate spatial signatures for multiple areas and group them into similar clusters based on their signatures [@Nowosad2021PatternbasedIdentification].

<!--
R, Python, and Julia have a wide range of packages for spatial operations, including *sf* [@Pebesma2018SimpleFeatures], or *terra* [@Hijmans2024TerraSpatial] in R, *GeoPandas* [@Jordahl2020GeopandasPython], or *rasterio* [@Gillies2013RasterioGeospatial] in Python, and *GeoStats.jl* [@Hoffimann2018GeoStatsjlHighperformance], or *GeoInterface.jl* [@JuliaGeoandcontributors2024GeoInterfaceJulia] in Julia.
These packages together with additional statistical and machine learning libraries allow for a wide range of spatial operations.
Additionally, the *motif* R package [@Nowosad2021MotifOpensource] provides a set of functions for spatial pattern operations, such as comparing, searching, or grouping spatial patterns customized for the use in landscape ecology.
-->

>>>>

@@

@@

## Vector metrics

@@

@@

Spatial patterns can be analyzed through a range of computational operations, e.g., comparing, searching, or grouping.
These operations are based on spatial signatures (multi-numerical representations of landscape pattern) and dissimilarity measures (functions that quantify differences between the signatures), and can be calculated for different areas or for the same area at two different moments in time.
Comparing spatial signatures is often used to analyze landscape dynamics, e.g., to detect changes in landscape structure over time [@Duncan2023MonitoringSpatial].
<!-- R1: As defined here, it looks to me that this operation is not necessarily linked to a query area, but rather to a spatial signature which can be directly specified or extracted from a query area -->
Furthermore, signature-based searches can be used to compare the spatial signature of a focal area to the spatial signatures of multiple other areas.
This allows for the identification of areas with similar signatures compared to the focal area, e.g., areas with similar environmental conditions [@Jasiewicz2014LandscapeSimilarity].

Additionally, it is possible to calculate spatial signatures for multiple areas and group them to similar clusters based on their signatures [@Nowosad2021PatternbasedIdentification].

### 2024-10-18 Sections/06-simulations.qmd

Pattern-based approaches simulate landscapes without assuming any underlying abiotic or biotic processes (i.e., neutral landscape model) [@Gardner1987NeutralModels].
The earliest neutral models are based on percolation theory and assign LULC classes randomly to cells in the landscape [@Gardner1987NeutralModels] or hierarchical models that consider different spatial scales while assigning cell values [@ONeill1992HierarchicalNeutral].
Landscapes characterized by continuous values can be simulated by fractal models, such as Brownian motion [@Palmer1992CoexistenceSpecies].
Borrowing from computer graphics, more recent neutral landscape models make use of spectral synthesis (e.g., Perlin noise) [@Etherington2022PerlinNoise] or binary space partitioning [@Etherington2022BinarySpace].
Neutral landscape models are also able to simulate landscapes dominated by anthropogenic activities based on least-cost networks [@Etherington2024LeastcostNetwork].
<!-- R1: Note that there, although the methods in reference 146 effectively rely on target values optimization (i.e. the model aims to get as close as possible to target values), the methods in reference 147 are rather based on target value satisfaction (the model guarantees that target values are strictly satisfied, or guarantees that they cannot be satisfied) -->
In order to ensure realistic neutral landscapes several approaches exist that use comparisons with real landscapes [@Inkoom2017DesigningNeutral], target values optimization (target values are optimized as close as possible) [@vanStrien2016ImprovedNeutral], or target value satisfaction (target values are strictly satisfied, or cannot be satisfied) [@Justeau-Allaire2022FragmentedLandscape].
<!-- In order to make multiple neutral landscape model algorithms accessible, several software libraries are available including Python [NLMpy\; @Etherington2015NLMpyPython], R [NLMR\; @Sciaini2018NLMRLandscapetools], Julia [NeutralLandscapes.jl\; @Poisot2023NeutralLandscapesJulia], or stand-alone software [GradientLand\; @Cambui2015GradientLandSoftware]. -->

>>>>

Pattern-based approaches simulate landscapes without assuming any underlying abiotic or biotic processes (i.e., neutral landscape model) [@Gardner1987NeutralModels].
The earliest neutral models are based on percolation theory and assign LULC classes randomly to cells in the landscape [@Gardner1987NeutralModels] or hierarchical models that consider different spatial scales while assigning cell values [@ONeill1992HierarchicalNeutral].
Landscapes characterized by continuous values can be simulated by fractal models, such as Brownian motion [@Palmer1992CoexistenceSpecies].
Borrowing from computer graphics, more recent neutral landscape models make use of spectral synthesis (e.g., Perlin noise) [@Etherington2022PerlinNoise] or binary space partitioning [@Etherington2022BinarySpace].
Neutral landscape models are also able to simulate landscapes dominated by anthropogenic activities based on least-cost networks [@Etherington2024LeastcostNetwork].
<!-- R1: Note that there, although the methods in reference 146 effectively rely on target values optimization (i.e. the model aims to get as close as possible to target values), the methods in reference 147 are rather based on target value satisfaction (the model guarantees that target values are strictly satisfied, or guarantees that they cannot be satisfied) -->
In order to ensure realistic neutral landscapes several approaches exist that use comparisons with real landscapes [@Inkoom2017DesigningNeutral], target values optimization (target values are optimized as close as possible) [@vanStrien2016ImprovedNeutral], or target value satisfaction (target values are strictly satisfied, or cannot be satisfied) [@Justeau-Allaire2022FragmentedLandscape].
